---
title: "Functions in R"
subtitle: "Basics - and some more - in R functions"
author: "Carmen Garro & Sebastian Ramirez-Ruiz"
date: "(Fall 2024) Introduction to Data Science"
output: 
    rmdformats::robobook:
    toc: TRUE
    df_print: paged
    number_sections: FALSE
    highlight: tango
    toc_depth: 3
    toc_float: true
    self_contained: false
---

```{=html}
<style>
.h1,h2,h3 {
color:#2f1a61;
}

.subtitle, section.normal {
color:#291854;
}

.title {
color:#cc0065;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Welcome! `r emo::ji("wave")`

Let's take a deeper look into functionality in R

---

# Important Concepts

## What is an R function?

* Functions in R are objects, just as data frames, vectors and lists are.
* Functions allow us to group code lines that serve a purpose and make them reusable.
* Functions are the abstract form of how a the code works.
* They can be broken down into 3 components: arguments, body and environment.

## Syntax and Components of a Function

A function in R will have the following structure:

```
first_function <- function(arguments){
  body of function
  }
```

## Components of a function

1. **Arguments**: what are the inputs that your function will work with. Called with *formals(function)*.
2. **Body**: this is the code inside the function. What you want to do with the inputs and define outputs. Called with *body(function)*
3. **Environment**: the data structure where your function finds the values associated with the inputs. *environment(function)*

```{r}
# function of summation
f_1 <- function(x,y){
  # sum of inputs
  result = x+y
  return(result)
}

formals(f_1)
# returns x and y

body(f_1)
# returns code inside of {}

environment(f_1)
# where the function is looking for x and y
```
---

# Types of Functions `r emo::ji("alien")`

## Primitive Functions

These are the functions present in baseR and don't require any package loading or function construction. They directly call C code to run, and are the exception to the rule of the three components.

All of these are primitive functions:

* sum()
* mean()
* sqrt()
* min()
* max()

```{r}
sum
```
They can be either builtin or special types. Closure type function are those that capture their environments.

```{r}
typeof(sum)
typeof(`[`)
typeof(f_1)
```

```{r}
formals(sum)
body(sum)
environment(sum)
```
## First Class Functions

Functions you create or load in R are objects, this property is called "First-Class Functions".  

You can name your functions or you can create anonymous ones, depending on your objective. Remember purrr's anonymous functions?

```{r}
purrr::map_int(1:3, ~.x + 1)
```
---

# Building Functions `r emo::ji("pencil")`

## Basic Functions in R

Previously we saw how the basic syntax of a function looks. We can use it as a guide to build our own functions in R.

When naming functions, a good practice to follow is to use **verbs** for the desired protocol to follow. Another good practice is to names them in an intuitive manner like "roll_the_dice". The most used arguments in basic functions are `x` and `y`.

Both baseR functions and arithmetic can be used to manipulate values.

The `...` is used to add non-specified arguments to the function.

```{r}
summation <- function(x,y, ...){
  result = sum(x,y,...)
  return(result)
}

summation(1,2,3,4,5,c(1,2,3,4))
```

We can also specify default values to the arguments of a function:

```{r}
# create dice simulation, and specify number of times to roll the dice
roll_the_dice <- function(n_times = 1){
  result = sample(1:6, n_times, replace = TRUE)
  return(paste("Rolling the dice",as.character(n_times),"times:",paste0(result, collapse = "-")))
}

roll_the_dice()
roll_the_dice(5)
```

## Conditional Statements in Functions

`if statements` allow for condition handling in functions.

Syntax
```
if(condition){
  what you want to happen if condition is true
} else {
  if condition is not true then do this
}
```


```{r}
roll_the_dice <- function(n_times = 1){
  if(!is.numeric(n_times)){
    print("Noup, I need a numerical input to work. Try again!")
  } else {
    result = sample(1:6, n_times, replace = TRUE)
  return(paste("Rolling the dice",as.character(n_times),"times:",paste0(result, collapse = "-")))
  }
}


roll_the_dice("Hallo")
```
## Error Handling

You can use if statements to handle errors, like in the last function. Let's see some other ways. You can use `stop()` to tell R what to output in case of an error where it cannot evaluate the function. 

```{r}
roll_the_dice <- function(n_times = 1){
  stop("Really, I need a number as an input")
  result = sample(1:6, n_times, replace = TRUE)
  return(paste("Rolling the dice",as.character(n_times),"times:",paste0(result, collapse = "-")))
}

# this produces an error with the desired message
# roll_the_dice(list(2))
```

Sometimes you want your function to be flexible and ignore or skip errors. You can use try() for this.

```{r}
#purrr::map(c(1,2,"three"), roll_the_dice)
```


```{r}
roll_the_dice <- function(n_times = 1){
  try(sample(1:6, n_times, replace = TRUE))
}

#purrr::map(c(1,2,"three"), roll_the_dice)
```
`try()` does not allow assigning names within it. If you want to do something further than ignoring the error you can use `tryCatch()`

```
f_1 <- function(name_input){
  tryCatch({your_function},
           error = function(e)
                    {e})
}
```

```{r}
roll_the_dice <- function(n_times = 1){
  tryCatch({
    result <- sample(1:6, n_times, replace = TRUE)
    return(paste("Rolling the dice", as.character(n_times),"times:",paste(result, collapse = "-")))},
    error = function(e){
      return(paste(paste0("'",as.character(n_times),"'"), "is not a numeric input..."))
    }
    )
}

purrr::map(c(1,2,"three"), roll_the_dice)
```
## Creating More Complex Functions

We can also add some elements we have worked with in past sessions. Let's create a function that creates possible variations of how a person can be called in documents.

This function works with names that have one first name and one last name, e.g. "Wael Sawan", CEO of Shell.

Possible variations:

Wael Sawal
W. Sawal
W Sawal
Wael S.
Wael S
W.S.

- Pseudocode - writing the steps I want my function to make in paper

first: split_name into first and last

construct variations:
complete  name
include first capital letter of name and complete last name
include first capital letter of name + . + complete last name
...

store them in a vector


```{r}
strsplit("Wael Sawan", " ")[[1]]
```



```{r}
generate_name_variations <- function(full_name) {
  tryCatch({
    # Split the name into first and last name
    name_parts <- strsplit(full_name, " ")[[1]]
    
    # Ensure that the name has exactly two parts
    if (length(name_parts) != 2) {
      stop("The input must be a two-word name.")
    }
    
    first_name <- name_parts[1]
    last_name <- name_parts[2]
    first_initial <- substr(first_name, 1, 1)
    last_initial <- substr(last_name, 1, 1)
    
    # Create the variations
    variations <- c(
      paste0(first_initial, ".", last_initial, "."),
      paste0(first_initial, ".", " ", last_initial, "."),
      paste0(first_initial, ".", " ", last_name),
      paste0(first_initial, " ", last_name),
      paste0(first_name, " ", last_name),
      paste0(first_name, " ", last_initial, "."),
      paste0(first_name, " ", last_initial)
    )
    
    return(variations)
  }, error = function(e) {
    # Handle the error and return a message
    return(paste("Error:", e$message))
  })
}

# Apply the function to a vector of names, using purrr::map or lapply
names_vector <- c("Wael Sawan", "John", "", "Alice Johnson", "NotAName")
purrr::map(names_vector, generate_name_variations)
```
Let's simplify this to show no output and create a regex pattern:

```{r}
generate_name_variations <- function(full_name) {
  tryCatch({
    # Split the name into first and last name
    name_parts <- strsplit(full_name, " ")[[1]]
    
    # Ensure that the name has exactly two parts
    if (length(name_parts) != 2) {
      e = ""
    } else
    {
    first_name <- name_parts[1]
    last_name <- name_parts[2]
    first_initial <- substr(first_name, 1, 1)
    last_initial <- substr(last_name, 1, 1)
    
    # Create the variations
    variations <- c(
      paste0(first_initial, ".", last_initial, "."),
      paste0(first_initial, ".", " ", last_initial, "."),
      paste0(first_initial, ".", " ", last_name),
      paste0(first_initial, ".", last_name),
      paste0(first_initial, " ", last_name),
      paste0(first_name, " ", last_name),
      paste0(first_name, " ", last_initial, "."),
      paste0(first_name, " ", last_initial),
      last_name
    )
    
    return(paste(variations, collapse = "|"))}
  }, error = function(e) {
    # return empty
    e
  })
}

# Apply the function to a vector of names, using purrr::map or lapply
names_vector <- c("Wael Sawan", "John", "", "Alice Johnson", "NotAName")
purrr::map(names_vector, generate_name_variations)
```

We can use the function to create variations of a list of names and create a string to match in a text. Let's use the names of the [Alabama US congress members](https://www.house.gov/representatives).

```{r}
al_cong <- c("Jerry Carl","Barry Moore","Miker Rogers","Robert Aderholt","Dale Strong","Gary Palmer","Sewell Terri")

texts_to_look <-c("J Carl is a congress member of the US","Robert A. and Strong are frequently seen discussing laws", "G.P. is from Alabama")

# Function to check if any variations of the name are mentioned in the texts
check_name_in_texts <- function(name, texts) {
  # Generate the name variations
  variations <- generate_name_variations(name)
  
  # Create a logical vector for each text, checking if any variation is found
  sapply(texts, function(text) {
    any(sapply(variations, function(variation) grepl(variation, text, ignore.case = TRUE)))
  })
}

# Create a data frame with names as rows and texts as columns
result_matrix <- sapply(al_cong, function(name) check_name_in_texts(name, texts_to_look))
result_df <- as.data.frame(t(result_matrix))

# Name the columns as text1, text2, etc.
colnames(result_df) <- paste0("text", seq_along(texts_to_look))

# Add the names of congress members as a column
result_df <- cbind(Name = al_cong, result_df)

# View the result
result_df <- as.data.frame(result_df)
row.names(result_df) <- NULL

result_df
  
```



---

References:

* Hadley Wickham. **Advanced R** Available at: https://adv-r.hadley.nz/functions.html#functionshttps://adv-r.hadley.nz/functions.html#functions
* Introduction to Data Science. Session 03 Tidyverse and Function. (2024). Hertie School. https://github.com/intro-to-data-science-24/labs/tree/main/session-03-tidyverse-functions
* ChatGPT was used to make code more efficient

